Landform boundary effects on Holocene forager landscape use in arid South Australia

### 2017-01-23 analysis/paper/paper.Rmd
initial

###########################################################
# checking P&M's edits from 14 May 2012
# clear the workspace
rm(list = ls())
# load data
load("E:/My Documents/My Papers/Journal of Field Archaeology/.RData")
setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology")

@@

ODX_allsites_2010 <- read.dbf("Join_Extract_ArcSite.dbf")

@@

setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology")

@@

ODX_TPs_2010 <- read.dbf("CLipped_ODO_terrain_patterns_2010_Ja.dbf")

@@

install.packages("ggplot2")
library(ggplot2)
ggplot(ODX_TP_areas, aes(reorder(GREGLANDF, prop), prop)) + 
  geom_bar(fill = "white", colour = "black") +
  theme_bw(base_size = 17) +
  opts(axis.text.x = theme_text(size = 17)) + 
  opts(axis.text.y = theme_text(size = 17)) +
  opts(axis.title.x = theme_text(size = 17)) + xlab("Terrain pattern") +
  opts(axis.title.y = theme_text(size = 17, angle = 90)) + ylab("Percentage of study area")
# check of sites per sq km for each TP
sites.per.sqkm <- merge(sites.per.tp, ODX_TP_areas, by.x = "Var1", by.y = "GREGLANDF")
sites.per.sqkm$sites.per <- with(sites.per.sqkm, Freq/V1*1000*1000)
# check of artefacts per sq km for each TP
artefacts.per.sqkm <- merge(artefacts.per.tp, ODX_TP_areas)
artefacts.per.sqkm$artefacts.per <- with(artefacts.per.sqkm, artefacts/V1*1000*1000)

# For looking a distance values from features in ArcMap
# First, polygon features have to be converted to line features.data management tools, features, feature to line.
# 1.  Open spatial analyst extension
# 2.	Define extent (LTA_areas is a good one)
# 3.	Spatial analyst - distance- straight line
# 4.	Then go to ArcToolbox - Spatial analyst tools - extraction 
# 5.	'Extract values to points'
# 6. Then arctoolbox, spatial analysis tools, extraction, extract values to points. point layer is sites, values come from raster, then RASTERVALU is distance field. 
# 7 .Then join to TP polygon to get TP values in sites dbf.
setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology")
library(foreign)
ODX_allsites_2010 <- read.dbf("Join_Extract_ArcSite.dbf")
ODX_allsites_2010 <- ODX_allsites_2010[ !is.na(ODX_allsites_2010$GREGLANDF), ]
ODX_allsites_2010.s <- subset(ODX_allsites_2010, GREGLANDF=='A4' | GREGLANDF=='Czs4' | GREGLANDF=='Q4' | GREGLANDF=='K4' | GREGLANDF=='Czs5' | GREGLANDF=='Q5' | GREGLANDF=='Q6')
ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF)
# order the levels of GREGLANDF for plotting
# this is the desired order: - A4, Czs4, Q4, K4, Czs5, Q5, Q6
levels(ODX_allsites_2010.s$GREGLANDF)
ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF, 
  levels(ODX_allsites_2010.s$GREGLANDF)[c(1, 2, 5, 4, 3, 6, 7)])
# plot
install.packages("ggplot2")
library(ggplot2)
ggplot(ODX_allsites_2010.s, aes(GREGLANDF, RASTERVALU)) + 
  geom_boxplot() + 
  scale_y_log10(breaks = seq(0,1400,200)) + 
  theme_bw(base_size = 17) +
  opts(axis.text.x = theme_text(size = 17)) + 
  opts(axis.text.y = theme_text(size = 12)) +
  opts(axis.title.x = theme_text(size = 17)) + xlab("Terrain pattern") +
  opts(axis.title.y = theme_text(size = 17, angle = 90)) + ylab("Distance of site to boundary (m)")

# anova to see if sig diff between distances...
fit <- aov(data = ODX_allsites_2010.s, RASTERVALU ~ GREGLANDF)
plot(TukeyHSD(fit))
# plot anova
install.packages("granovaGG", dependencies = TRUE)
library("granovaGG")
granovagg.1w(data = ODX_allsites_2010.s$RASTERVALU, group = ODX_allsites_2010.s$GREGLANDF)
install.packages("granova")
library("granova")
granova.1w(data = ODX_allsites_2010.s$RASTERVALU, group = ODX_allsites_2010.s$GREGLANDF)

setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology\\outer_buffers")
# start with OUTER buffers
bd_files <- list.files()
# list in, dataframe out
library(plyr)
library(foreign)
bd.data <- ldply(bd_files, read.dbf)
bd.data$SiteDens <- with(bd.data, Count_/buf_area * (1000*1000))
# exclude buffers with less than n sites since they 
# skew the density calculations 
n = 15
bd.data <- bd.data[bd.data$Count_ > n, ]
unique(bd.data$BUFF_DIST)
bd.data$Buf <- as.factor(with(bd.data, ifelse(BUFF_DIST == 100, '000-100', ifelse(BUFF_DIST == 200, '100-200', ifelse(BUFF_DIST == 300, '200-300', ifelse(BUFF_DIST == 400, '300-400', ifelse(BUFF_DIST == 500, '400-500','NA')))))))

@@

# now check the boundary densities for INNER buffers
setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology\\inner_buffers")
inbd.files <- list.files()
# list in, dataframe out
library(plyr)
library(foreign)
inbd.data <- ldply(inbd.files, read.dbf)
inbd.data$SiteDens <- with(inbd.data, Join_Count/buf_area * (1000*1000))
inbd.data <- inbd.data[inbd.data$Join_Count > n, ]
unique(inbd.data$BUFF_DIST)
inbd.data$Buf <- as.factor(with( inbd.data, ifelse(is.na(BUFF_DIST), '-100-000', ifelse(BUFF_DIST == -100, '-200-100', ifelse(BUFF_DIST == -200, '-300-200', ifelse(BUFF_DIST == -300, '-400-300', ifelse(BUFF_DIST == -400, '-500-400','NA')))))))

@@

# combine inner with outer buffers
inner <- subset(inbd.data, GREGLANDF =='Czs5' | GREGLANDF =='Czs4')[,c('GREGLANDF','SiteDens','Buf')]
outer <- subset(bd.data, GREGLANDF =='Czs5' | GREGLANDF =='Czs4')  [,c('GREGLANDF','SiteDens','Buf')]
# colnames(outer)[2] <-  'mean_SiteDens'
inner_outer <- na.omit(rbind(inner, outer))
# put the boundary distances in order
levels(inner_outer$Buf)
inner_outer$Buf <- factor(inner_outer$Buf, 
  levels(inner_outer$Buf)[c(5,4,3,2,1,6,7,8,9,10)])
# remove rows with zero for SiteDens
inner_outer <- inner_outer[inner_outer[, 'SiteDens']!=0, ]  
# plot inner and outer together
library(ggplot2)
library(grid)
ggplot(inner_outer,  aes(Buf, SiteDens)) + 
  geom_point(size = 2) + facet_grid(GREGLANDF ~ .) + 
  scale_y_log10() + 
  geom_smooth(aes(group = 1), fill = "grey80", colour = "grey") +
  theme_bw() + 
  opts(axis.title.x = theme_text(vjust = -0.5, size = 15)) + 
  opts(axis.title.y=theme_text(size = 15, angle=90)) + 
  opts(plot.margin = unit(c(1,1,2,2), "lines")) + 
  xlab("Distance from boundary in meters (negative values are interior distances, positive values are exterior distances)") + 
  ylab("Sites per sq km")

@@

library(coin) # library needed for more permutation tests
# We'll use Monte Carlo simulations to get the p value
oneway_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs5',], SiteDens ~ Buf, distribution=approximate(B=10000))
# this is a test of whether 
# two or more samples from normal distributions have the 
# same means, variances not assumed equal
kruskal_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs5',], SiteDens ~ Buf, distribution=approximate(B=10000))
# yes for Csz5, no for Csz4

@@

# plot
library(ggplot2)
ggplot(bound.m, aes(dist,value, colour = variable)) + geom_line(aes(group = variable))

@@

# plot scales
library(ggplot2)
ggplot(bound.sc.m, aes(x=bound.sc.m[,1], y=bound.sc.m[,3])) + 
  geom_line(aes(group = bound.sc.m[,2], linetype = bound.sc.m[,2]), size = 1)  +
  theme_bw() +
  opts(axis.text.x = theme_text(size = 17)) + 
  opts(axis.text.y = theme_text(size = 17)) +
  opts(axis.title.x = theme_text(size = 17)) +
  opts(axis.title.y = theme_text(size = 17, angle = 90)) + 
  xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
  ylab("z-score of site density in buffer area") +
  scale_linetype_manual(name="Terrain Pattern", values = c(1,2) ) +
  opts(legend.text=theme_text(size=14)) +
  opts(legend.title=theme_text(size=14))

>>>>

###########################################################
# checking P&M's edits from 14 May 2012
#
# clear the workspace
rm(list = ls())
#
# load data
load("../data/.RData")
#

@@

ODX_allsites_2010 <- read.dbf("../data/Join_Extract_ArcSite.dbf")

@@

@@

ODX_TPs_2010 <- read.dbf("../data/CLipped_ODO_terrain_patterns_2010_Ja.dbf")

@@

library(ggplot2)
ggplot(ODX_TP_areas, aes(reorder(GREGLANDF, prop), prop)) + 
  geom_bar(stat = "identity",
           fill = "white", colour = "black") +
  theme_bw(base_size = 17) +
  theme(axis.text.x = element_text(size = 17),
        axis.text.y = element_text(size = 17),
        axis.title.x = element_text(size = 17),
        axis.title.y = element_text(size = 17, angle = 90)) + 
  ylab("Percentage of study area") +
  xlab("Terrain pattern") 
#
# check of sites per sq km for each TP
sites.per.sqkm <- merge(sites.per.tp, ODX_TP_areas, by.x = "Var1", by.y = "GREGLANDF")
sites.per.sqkm$sites.per <- with(sites.per.sqkm, Freq/V1*1000*1000)
#
# check of artefacts per sq km for each TP
artefacts.per.sqkm <- merge(artefacts.per.tp, ODX_TP_areas)
artefacts.per.sqkm$artefacts.per <- with(artefacts.per.sqkm, artefacts/V1*1000*1000)

# For looking a distance values from features in ArcMap
# First, polygon features have to be converted to line features.data management tools, features, feature to line.
# 1.  Open spatial analyst extension
# 2.	Define extent (LTA_areas is a good one)
# 3.	Spatial analyst - distance- straight line
# 4.	Then go to ArcToolbox - Spatial analyst tools - extraction 
# 5.	'Extract values to points'
# 6. Then arctoolbox, spatial analysis tools, extraction, extract values to points. point layer is sites, values come from raster, then RASTERVALU is distance field. 
# 7 .Then join to TP polygon to get TP values in sites dbf.

library(foreign)
ODX_allsites_2010 <- read.dbf("../data/Join_Extract_ArcSite.dbf")
ODX_allsites_2010 <- ODX_allsites_2010[ !is.na(ODX_allsites_2010$GREGLANDF), ]
ODX_allsites_2010.s <- subset(ODX_allsites_2010, GREGLANDF=='A4' | GREGLANDF=='Czs4' | GREGLANDF=='Q4' | GREGLANDF=='K4' | GREGLANDF=='Czs5' | GREGLANDF=='Q5' | GREGLANDF=='Q6')
ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF)
# order the levels of GREGLANDF for plotting
# this is the desired order: - A4, Czs4, Q4, K4, Czs5, Q5, Q6
levels(ODX_allsites_2010.s$GREGLANDF)
ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF, 
  levels(ODX_allsites_2010.s$GREGLANDF)[c(1, 2, 5, 4, 3, 6, 7)])

# plot
library(ggplot2)
ggplot(ODX_allsites_2010.s, aes(GREGLANDF, RASTERVALU)) + 
  geom_boxplot() + 
  scale_y_log10(breaks = seq(0,1400,200)) + 
  theme_bw(base_size = 17) +
  theme(axis.text.x =  element_text(size = 17),
        axis.text.y =  element_text(size = 12),
        axis.title.x = element_text(size = 17),
        axis.title.y = element_text(size = 17, 
                                    angle = 90))  +
  xlab("Terrain pattern") +
  ylab("Distance of site to boundary (m)")

# anova to see if sig diff between distances...
fit <- aov(data = ODX_allsites_2010.s, RASTERVALU ~ GREGLANDF)
plot(TukeyHSD(fit))

# plot anova
# install.packages("granovaGG", dependencies = TRUE)
library("granovaGG")
granovagg.1w(data = ODX_allsites_2010.s$RASTERVALU, group = ODX_allsites_2010.s$GREGLANDF)
# install.packages("granova")
library("granova")
granova.1w(data = ODX_allsites_2010.s$RASTERVALU, group = ODX_allsites_2010.s$GREGLANDF)

# start with OUTER buffers
bd_files <- list.files("../data/outer_buffers", full.names = TRUE)
# list in, dataframe out
library(plyr)
library(foreign)
bd.data <- ldply(bd_files, read.dbf)
bd.data$SiteDens <- with(bd.data, Count_/buf_area * (1000*1000))
# exclude buffers with less than n sites since they 
# skew the density calculations 
n = 15
bd.data <- bd.data[bd.data$Count_ > n, ]
unique(bd.data$BUFF_DIST)
bd.data$Buf <- as.factor(with(bd.data, ifelse(BUFF_DIST == 100, '000-100', ifelse(BUFF_DIST == 200, '100-200', ifelse(BUFF_DIST == 300, '200-300', ifelse(BUFF_DIST == 400, '300-400', ifelse(BUFF_DIST == 500, '400-500','NA')))))))

@@

# now check the boundary densities for INNER buffers
inbd.files <- list.files("../data/inner_buffers", full.names = TRUE)
# list in, dataframe out
library(plyr)
library(foreign)
inbd.data <- ldply(inbd.files, read.dbf)
inbd.data$SiteDens <- with(inbd.data, Join_Count/buf_area * (1000*1000))
inbd.data <- inbd.data[inbd.data$Join_Count > n, ]
unique(inbd.data$BUFF_DIST)
inbd.data$Buf <- as.factor(with( inbd.data, ifelse(is.na(BUFF_DIST), '-100-000', ifelse(BUFF_DIST == -100, '-200-100', ifelse(BUFF_DIST == -200, '-300-200', ifelse(BUFF_DIST == -300, '-400-300', ifelse(BUFF_DIST == -400, '-500-400','NA')))))))

@@

# combine inner with outer buffers
inner <- subset(inbd.data, GREGLANDF =='Czs5' | GREGLANDF =='Czs4')[,c('GREGLANDF','SiteDens','Buf')]
outer <- subset(bd.data, GREGLANDF =='Czs5' | GREGLANDF =='Czs4')  [,c('GREGLANDF','SiteDens','Buf')]
# colnames(outer)[2] <-  'mean_SiteDens'
inner_outer <- na.omit(rbind(inner, outer))
# put the boundary distances in order
levels(inner_outer$Buf)
inner_outer$Buf <- factor(inner_outer$Buf, 
  levels(inner_outer$Buf)[c(5,4,3,2,1,6,7,8,9,10)])
# remove rows with zero for SiteDens
inner_outer <- inner_outer[inner_outer[, 'SiteDens']!=0, ]  
# plot inner and outer together
library(ggplot2)
library(grid)
library(scales)
ggplot(inner_outer,  
       aes(Buf, 
           SiteDens)) + 
  geom_point(size = 2) + 
  facet_grid(GREGLANDF ~ .) + 
  scale_y_log10() + 
  geom_smooth(aes(group = 1), 
              fill = "grey80", 
              colour = "grey") +
  theme_bw() + 
  theme(
  axis.title.x = element_text(vjust = -0.5, 
                              size = 15),
  axis.title.y=  element_text(size = 15, 
                              angle=90), 
  plot.margin = unit(c(1,1,2,2), "lines")) + 
  xlab("Distance from boundary in meters \n(negative values are interior distances, \npositive values are exterior distances)") + 
  ylab("Sites per sq km")

@@

library(coin) # library needed for more permutation tests
# We'll use Monte Carlo simulations to get the p value
oneway_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs5',], SiteDens ~ Buf, distribution=approximate(B=10000))
oneway_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs4',], SiteDens ~ Buf, distribution=approximate(B=10000))
# this is a test of whether 
# two or more samples from normal distributions have the 
# same means, variances not assumed equal
kruskal_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs5',], SiteDens ~ Buf, distribution=approximate(B=10000))
kruskal_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs4',], SiteDens ~ Buf, distribution=approximate(B=10000))
# yes for Csz5, no for Csz4

@@

# plot
library(ggplot2)
ggplot(bound.m, 
       aes(dist,
           value, 
           colour = variable)) + 
  geom_line(aes(group = variable))

@@

# plot scales
library(ggplot2)
ggplot(bound.sc.m, 
       aes(x=bound.sc.m[,1], 
           y=bound.sc.m[,3])) + 
  geom_line(aes(group = bound.sc.m[,2], 
                linetype = bound.sc.m[,2]), 
            size = 1)  +
  theme_bw() +
  theme(
  axis.text.x =  element_text(size = 17), 
  axis.text.y =  element_text(size = 17),
  axis.title.x = element_text(size = 17),
  axis.title.y = element_text(size = 17,
                              angle = 90),
  legend.text =  element_text(size=14),
  legend.title = element_text(size=14)) + 
  xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
  ylab("z-score of site density in buffer area") +
  scale_linetype_manual(name="Terrain Pattern", values = c(1,2) )
